---
title: "Intro to AutoTuner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE}
library(Autotuner)
library(devtools)
if(!require("mmetspData")) {
    install_github("crmclean/mmetspData")    
    library(mmetspData)
}
```

# Introduction

AutoTuner is a parameter tuning algorithm for XCMS, mzMine2, and other 
metabolomics data processing softwares. Using statistical inference, AutoTuner
quickly finds estimates for nine distinct parameters. This guide provides
an interactive example of how to use AutoTuner.

Here is the general description of AutoTuner:

```{r fig.width=8, fig.height=5,echo=FALSE}
library(png)
library(grid)
img <- readPNG("autotuner_schematic_final.png")
 grid.raster(img)
```

## Set up:

### Input - Raw Data

AutoTuner is designed to work directly with raw mass spectral data that has 
been processed by msconvert. So far file types .mzML, .mzXML, .msData, and
.netCDF have been tested to work well. 

```{r loading mass spec data}
mmetspFiles <- c(system.file("mzMLs/mtab_mmetsp_ft_120815_24.mzML", package = "mmetspData"),
                 system.file("mzMLs/mtab_mmetsp_ft_120815_25.mzML", package = "mmetspData"),
                 system.file("mzMLs/mtab_mmetsp_ft_120815_26.mzML", package = "mmetspData"))
```

Here are the filetypes that will be used within this tutorial:

```{r filetype}
print(basename(mmetspFiles))
```

### Input - Metadata

AutoTuner also requires a metadata spreadsheet that has at least two columns 
in order to derive estimates. One column should contain string matches to all
the raw data files that will be processed. The second should contain information
on the experimental factor each sample belongs to. If you are using Shiny to 
run AutoTuner, the file has to have .csv formal. Future implementations of AutoTuner
will feature a broader suite of acceptable file types. 

```{r loading in metadata}
metadata <- read.csv(system.file("mmetsp_metadata.csv", package = "mmetspData"),
                    stringsAsFactors = F)

metadata <- metadata[metadata$File.Name %in% 
                         sub(pattern = ".mzML", "",basename(mmetspFiles)),]
```

This is what the metadata file should look like. In our case, the column 
matching the raw data files is called "File.Name", while the one with 
experimental factor information is called "Sample.Type".

```{r}
print(metadata)
```

## Setting up AutoTuner

### Creating AutoTuner Object

AutoTuner first requires that user create an AutoTuner object. All future
computations will be contained within the object. 

The file_col argument corresponds to the string column of the metadata that 
matches raw data samples by name. The factorCol argument corresponds to the 
specific factor column. 

```{r}
Autotuner <- createAutotuner(mmetspFiles,
                             metadata,
                             file_col = "File.Name",
                             factorCol = "Sample.Type")
```

## Part 1: Total Ion Current Peak Identification

The first part of AutoTuner involves the identification of peaks within the
total ion current (TIC) of the samples loaded up into AutoTuner. These regions will
be important later to estimate parameters from the raw data since it is assumed
that they contain a greater number of real chemical measurements. 

### Sliding Window Analysis

To do this, first comes the sliding window analysis. A sliding windown analysis 
is a simple time series analysis algorithm used to identify peaks within a time 
trace essentially using a moving average. More on sliding window analyses can be
found [here](https://en.wikipedia.org/wiki/Moving_average).

The goal of this analysis is to find peaks within the TIC. The user should 
prioritize finding where peaks *start* rather than caturing the entire peak 
bound. Downstream steps actually do a better job of estimating what the peaks 
bounds should be. 

The user should play with the lag, threshold, and influence parameters to 
perform the sliding window analysis. Here is what they represent relative to
chromatography:

Lag - The number of chromatographic scan points used to test if next point is
significance. 
Threshold - A numerical constant representing how many times greater the 
intensity of an adjacent scan has to be from the scans in the sliding window
to be considered significant. 
Influence - A numerical factor used to scale the magnitude of a significant scan
once it has been added to the sliding window. 

```{r}
lag <- 25
threshold<- 3.1
influence <- 0.1
signals <- lapply(getAutoIntensity(Autotuner), 
                 ThresholdingAlgo, lag, threshold, influence)
```

The output of the sliding window can be displayed with the plot_signals
function:

```{r}
plot_signals(Autotuner, 
             threshold, 
             ## index for which data files should be displayed
             sample_index = 1:3, 
             signals = signals)
rm(lag, influence, threshold)
```

The figure above has two components:

Top Plot: The chromatotgraphic trace for each sample (solid line) along with the noise 
associated with each sample (dashed line). 
Bottom Plot: A signal plot used to indicate which chromatographic regions have significant
peaks.

The user should look for combinations of the three sliding window parameters that
return lots of small peaks within the plot below the chromatographic trace. 

Autotuner will expand each of these regions to obtain improved estimates on the
bounds within the isolatePeaks function below. The return_peaks arguement there
represents the number of peaks to return for parameter estimation. This number
is bounded by the total number of detected peaks by the sliding windown analysis
above, so seeing more narrow peaks within the signal plot is recommended. 

```{r}
Autotuner <- isolatePeaks(Autotuner, returned_peaks = 10, signals)
```

The peaks with expanded bounds returned from the isolatePeaks function can be 
rapidly checked visually using the plot_peaks function as shown below. 
The bounds should capture the correct ascention and descention points of each 
peak. If peak bounds are not satisfactory, the user should return to the 
sliding window analysis, and try a different conbination of the three parameters.

Remember, this whole process is only designed to isolate regions enriched in 
real features rather than find true peaks. The bounds don't need to be
completelly perfect. Its much more important that the bounds contain some kind 
of chromatographic peaks rather than less dynamic regions of the chromatographic
trace.

```{r}
for(i in 1:5) {
    plot_peaks(Autotuner = Autotuner, 
           boundary = 100, 
           peak = i)    
}
```

## Part 2: Parameter Extraction from Individual Extracted Ion Chromatograms

In order to estimate parameters from the raw data, the user should run the 
EICparams function as below. The massThreshold is an absolute mass error that
should be greater than the expected analytical capabilities of the mass analyzer.
This part of the analysis might take a few minutes if the data used is large 
(~ 100 Mb per sample).

If returnPpmPlots is True, AutoTuner will return plots showing how the ppm 
threshold was estimated within the current working directory running Autotuner. 
This can be used to evaluate the magnitude of the massThreshold parameter.

```{r}
## error with peak width estimation
## idea - filter things by mass. smaler masses are more likely to be random assosications
eicParamEsts <- EICparams(Autotuner = Autotuner, 
                          massThresh = .005, 
                          verbose = F,
                          returnPpmPlots = F,
                          useGap = T)
```

## Part 3: Returning Estimates

All that remains now is to get what the dataset estimates are.

```{r}
returnParams(eicParamEsts, Autotuner)
```

```{r}
sessionInfo()
```

